Seasonality of submesoscale dynamics in the Kuroshio Extension

### 2016-07-13 writeup/rocha_etal.tex
Dimitris' comments

\keypoints{\item  Upper-ocean submesoscale (10-100 km) turbulence and inertia-gravity waves
                  undergo strong seasonal cycles that are out-of-phase.
           \item  Submesoscale turbulence dominates the horizontal velocity and
                  sea-surface height variability in late winter/early spring.
            \item Submesoscale inertia-gravity waves dominate the horizontal velocity and
                   sea-surface height variability in late summer/early fall.
                  }

@@

Using the output of $1/24^\circ$ and $1/48^\circ$ global
numerical simulations with embedded tides, we show that inertia-gravity waves undergo
a strong seasonality near the surface in the Kuroshio Extension region.
Interestingly, the seasonal cycle of inertia-gravity waves is out-of-phase
with the seasonal cycle of submesoscale turbulence. Consistent with previous studies,
deep late-winter mixed layers are prone to
shallow baroclinic instabilities that are roughly in geostrophic balance
and flux energy upscale \citep{sasaki_etal2014,callies_etal2016},
driving a mild seasonal modulation of the mesoscales \citep{sasaki_etal2014,qiu_etal2014}.

@@

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
realistic numerical simulations. The outputs
analyzed here, LLC2160 (1/24$^\circ$) and LLC4320 (1/48$^\circ$),   are
forward Massachusetts Institute of Technology general circulation model (MITgcm)
numerical solutions on a LLC grid  \citep{forget_etal2015} with
90-vertical-levels. The coarser-resolution LLC simulation was
spun up from an ECCO2 adjoint-method state estimate, constrained to millions
of observations from 2009 through 2011. Both simulations were forced by
tides and 6-hourly surface fluxes. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the 3-month long output used by
\citet{rocha_etal2016}.  Details of the LLC simulations are provided in the supplemental material.

@@

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinity of the Kuroshio
Extension, where previous studies have shown strong mesoscale and submesoscale seasonality
\citep{sasaki_etal2014,qiu_etal2014}.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (Figure \ref{fig1}). The stratification
in this mesoscale-rich subtropical region undergoes a vigorous seasonal cycle: wintertime-enhanced
small-scale turbulence de-stratifies the upper ocean, yielding mixed layers
as deep as 300m. In late spring/early summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40m.
Fundamentally, the upper-ocean density structure is well-captured by both LLC simulations:
a comparison with Argo climatology shows that both simulations skillfully represent the Kuroshio
Extension stratification and its seasonal variability  (supplemental material).

@@

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early spring
(April 15) and fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions depict seasonality in vorticity: large values of
fine-grained vertical vorticity are observed in early spring with maximum
values as large as $4f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about $0.4f$. In early
fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained;
its local maximum and RMS are both smaller than $0.5f$.
Indeed, the vorticity and rate of strain are strongest in wintertime (Figure \ref{fig2}):
in both simulations, the RMS vorticity and strain rate are about twice as large in
late winter/early spring
than in late summer/early fall. Because the wintertime vorticity and strain rate
are dominated by the smallest scales in the flow (the KE spectra is shallower than
a $-3$ power-law in winter), increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ increases the wintertime RMS vorticity
and strain by about 40$\%$.

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.75\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square (RMS) of vorticity (a) and
  rate of strain (b), and horizontal divergence (c) in the LLC outputs and gridded AVISO data.}
  \label{fig2}
  \end{center}
\end{figure*}

@@

The seasonal cycle of the horizontal divergence, however, showcases the complexity
of the upper-ocean annual variability. If submesoscale eddies and fronts dominated
the near-surface variability all year, then the seasonal cycle of horizontal divergence,
vertical vorticity, and lateral strain rate would be in phase \citep[e.g.,][]{sasaki_etal2014}.
While there is a clear wintertime peak in divergence of daily-averaged velocity
(see dashed lines in figure \ref{fig2}c; RMS divergence $\sim0.22 f$ in the 1/48$^\circ$
simulation), the hourly fields show
a stronger enhancement of
lateral divergence in late summer/early fall (RMS divergence $\sim0.22 f$ the 1/48$^\circ$
simulation). Because the 1/48$^\circ$ simulation better resolves
smaller-scale submesoscale flows, a secondary RMS divergence peak in winter  is
nearly as strong as in summer. Submesoscale fronts and eddies evolve
relatively fast, and there is no clear
temporal and spatial scale separation between those motions and inertia-gravity waves
\citep{mcwilliams2016}:
daily-averaging the velocity fields efficiently suppresses the summertime horizontally
divergent flows, but also reduces the wintertime lateral divergence by about 50$\%$.
Figure \ref{fig2}c also shows that most of the lateral divergence is associated
with submesoscale flows: smoothing the velocity fields with a 100km-cut-off
suppresses more than 80$\%$ of the RMS divergence.

\begin{figure*}[ht]
  \begin{center}
    \includegraphics[width=.95\textwidth]{figs/fig3.pdf}
 \caption{Seasonal variation of joint probability distributions:  vorticity vs. strain rate (a through d),
          and vorticity vs. Laplacian of sea-surface height (e through h) in April (a, b,
          e, f) and October (c, d, g, h).
          Dashed lines in (a) through (d) represent one dimensional shear flow $\alpha = \pm\zeta$,
          characteristic of fronts. Dashed lines in (e) through
          (h) represent geostrophic flow $\zeta = \frac{g}{f}\nabla_h^2 \eta$.}
 \label{fig3}
 \end{center}
\end{figure*}

@@

The April vorticity-strain jPDF has a shape characteristic of submesoscale turbulence
 (Figures \ref{fig3}a-b). The alignment of vorticity and strain $\alpha \sim \pm\zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly, daily-averaged, although
 the vorticity skewness reduces from 1.4 to 1.13 from hourly to daily-averaged.
 The April results are characteristic of winter, indicating that wintertime submesoscale
 surface velocity is strongly dominated by submesoscale turbulence.
 The hourly velocity fields are significantly
ageostrophic as depicted by the jPDF of vorticity-Laplacian of SSH
winter (Figure \ref{fig3}e). Even in April, only the daily-averaged fields
are largely
in  geostrophic balance (the jPDF of daily-averaged vorticity vs. Laplacian of SSH
is an ellipse with large eccentricity and main axis tilted by 45$^\circ$; Figure \ref{fig3}f).

The October vorticity-strain jPDF shows much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
that persist in summer) and one with a very steep slope.
That the submesoscale dynamics in October are mainly unbalanced is clearly depicted
in the shape of jPDF of vorticity-Laplacian of SSH for hourly fields
winter (Figure \ref{fig3}g), which is an ellipse aligned in the vertical axis.
Of course daily-averaging the model suppresses the ageostrophic, super-inertial flows, and
the daily-averaged flow is essentially geostrophic as depicted by the 45$^\circ$-tilted
ellipse in the vorticity-Laplacian of SSH jPDF.

@@

To better quantify the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance.  Before calculating the spectra, time-mean and spatial linear
trends were removed, and the resulting fields were multiplied by a two-dimensional Hanning ``spectral window''.
The two-dimensional spectra were averaged azimuthly \citep[e.g., ][]{rocha_etal2016} .
We discuss only spectra for the 1/48$^\circ$ simulation, which extends the 1/24$^\circ$ simulation
towards smaller scales (supplemental material).

\begin{figure*}[ht]
  \begin{center}
    \includegraphics[width=1.\textwidth]{figs/fig4.pdf}
 \caption{Surface (horizontal) KE (a) and SSH variance wavenumber spectra (b)
 in the 1/48$^\circ$ simulation. Solid lines
 are spectra based on hourly snapshots, dashed lines are spectra based on daily-averaged
 fields.}
 \label{fig4}
 \end{center}
\end{figure*}

Figure \ref{fig4}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the April and October spectra based on hourly (the solid curves)
velocity snapshots are nearly indistinguishable from each other within 95$\%$ confidence level.
Consistent with the results of
\citet{rocha_etal2016}, who analyzed a 3-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig4}a). But this suppression is dramatic in October, when the inertia-gravity waves peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

@@

An important caveat is that this study focuses on a single patch of ocean in the
vicinity of the Kuroshio Extension. This may be a typical of mesoscale-rich
subtropical region, but it is unlikely to be
representative of other regions such as low eddy kinetic energy eastern boundary currents
and the middle of the subtropical gyre.  We plan to report on the geographic variability of submesoscale
 seasonality in a future study.

@@

% say something about relative energy in smoother vs. unsmoothed fields.
% also say something about comparison with observations.

@@

\begin{acknowledgments}
 William R. Young provided helpful feedback on the first draft. We thank the MITgcm community and our colleagues at the NASA Advanced
Supercomputing (NAS) Division for their awesome support.
This research was funded by NSF (OCE1357047) and NASA (NNX13AE44G,NNX13AE85G,NNX16AH67G).
The LLC outputs can be obtained from D. Menemenlis (menemenlis@jpl.nasa.gov). The altimeter products were produced by Ssalto/Duacs
and distributed by Aviso, with support from CNES (\texttt{http://www.aviso.altimetry.fr/duacs/}).
\end{acknowledgments}

>>>>

\keypoints{\item  Upper-ocean submesoscale (10-100 km) turbulence and inertia-gravity waves
                  undergo strong seasonal cycles that are out of phase.
           \item  Submesoscale turbulence dominates the horizontal velocity and
                  sea-surface height variability in late winter/early spring.
            \item Submesoscale inertia-gravity waves dominate the horizontal velocity and
                   sea-surface height variability in late summer/early fall.
                  }

@@

Using the output of $1/24^\circ$ and $1/48^\circ$ global
numerical simulations with embedded tides, we show that inertia-gravity waves undergo
a strong seasonality near the surface in the Kuroshio Extension region.
Interestingly, the seasonal cycle of inertia-gravity waves is out of phase
with the seasonal cycle of submesoscale turbulence. On one hand, consistent with previous studies,
deep late-winter mixed layers are prone to
shallow baroclinic instabilities that are roughly in geostrophic balance
and flux energy upscale \citep{sasaki_etal2014,callies_etal2016},
driving a mild seasonal modulation of the mesoscales \citep{sasaki_etal2014,qiu_etal2014}.

@@

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
realistic numerical simulations. The outputs
analyzed here, LLC2160 (1/24$^\circ$) and LLC4320 (1/48$^\circ$),   are
forward Massachusetts Institute of Technology general circulation model (MITgcm)
numerical solutions on a LLC grid  \citep{forget_etal2015} with
90-vertical-levels. The coarser-resolution LLC simulation was
spun up from an Estimating the Circulation and Climate of the Ocean, Phase II (ECCO2)
adjoint-method state estimate, constrained to millions
of observations from 2009 through 2011. Both simulations were forced by
tides and 6-hourly surface atmospheric fields. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the 3-month long output used by
\citet{rocha_etal2016}.  Details of the LLC simulations are provided in the supplemental material.

@@

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinity of the Kuroshio
Extension, where previous studies have shown strong mesoscale and submesoscale seasonality
\citep{sasaki_etal2014,qiu_etal2014}.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (Figure \ref{fig1}). The stratification
in this mesoscale-rich subtropical region undergoes a vigorous seasonal cycle: wintertime-enhanced
small-scale turbulence de-stratifies the upper ocean, yielding mixed layers
as deep as 300 m. In late spring/early summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40 m.
Fundamentally, the upper-ocean density structure is well-captured by both LLC simulations:
a comparison with Argo climatology shows that both simulations skillfully represent the Kuroshio
Extension stratification and its seasonal variability  (supplemental material).

@@

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early spring
(April 15) and fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions depict seasonality in vorticity: large values of
fine-grained vertical vorticity are observed in early spring with maximum
values as large as $4f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about $0.4f$. In early
fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained;
its local maximum and RMS are both smaller than $0.5f$.
Indeed, the vorticity and rate of strain are strongest in wintertime (Figure \ref{fig2}):
in both simulations, the RMS vorticity and strain rate are about twice as large in
late winter/early spring
than in late summer/early fall. Because the wintertime vorticity and strain rate
are dominated by the smallest scales in the flow (the KE spectra is shallower than
a $-3$ power law in winter), increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ increases the wintertime RMS vorticity
and strain by about 40$\%$.

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.75\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square (RMS) of (a) vorticity,
  (b) rate of strain, and (c) horizontal divergence in the LLC outputs and gridded AVISO data.}
  \label{fig2}
  \end{center}
\end{figure*}

@@

The seasonal cycle of the horizontal divergence, however, showcases the complexity
of the upper-ocean annual variability. If submesoscale eddies and fronts dominated
the near-surface variability all year, then the seasonal cycle of horizontal divergence,
vertical vorticity, and lateral strain rate would be in phase \citep[e.g.,][]{sasaki_etal2014}.
While there is a clear wintertime peak in divergence of daily-averaged velocity
(see dashed lines in figure \ref{fig2}c; RMS divergence $\sim0.1 f$ in the 1/48$^\circ$
simulation), the hourly fields show
a stronger enhancement of
lateral divergence in late summer/early fall (RMS divergence $\sim0.22 f$ the 1/48$^\circ$
simulation). Because the 1/48$^\circ$ simulation better resolves
smaller-scale submesoscale flows, a secondary RMS divergence peak in winter  is
nearly as strong as in summer. Submesoscale fronts and eddies evolve
relatively fast, and there is no clear
temporal and spatial scale separation between those motions and inertia-gravity waves
\citep{mcwilliams2016}:
daily-averaging the velocity fields efficiently suppresses the summertime horizontally
divergent flows, but also reduces the wintertime lateral divergence by about 50$\%$.
Figure \ref{fig2}c also shows that most of the lateral divergence is associated
with submesoscale flows: smoothing the velocity fields with a 100-km-cutoff
suppresses more than 80$\%$ of the RMS divergence.

\begin{figure*}[ht]
  \begin{center}
    \includegraphics[width=.95\textwidth]{figs/fig3.pdf}
 \caption{Seasonal variation of joint probability distributions:  vorticity vs.\ strain rate (a through d),
          and vorticity vs.\ Laplacian of sea-surface height (e through h) in April (a, b,
          e, f) and October (c, d, g, h).
          Dashed lines in (a) through (d) represent one dimensional shear flow $\alpha = \pm\zeta$,
          characteristic of fronts. Dashed lines in (e) through
          (h) represent geostrophic flow $\zeta = \frac{g}{f}\nabla_h^2 \eta$.}
 \label{fig3}
 \end{center}
\end{figure*}

@@

The April vorticity-strain jPDF has a shape characteristic of submesoscale turbulence
 (Figures \ref{fig3}a-b). The alignment of vorticity and strain $\alpha \sim \pm\zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly and daily-averaged fields, although
 the vorticity skewness reduces from 1.4 to 1.13 from hourly to daily-averaged.
 The April results are characteristic of winter, indicating that wintertime submesoscale
 surface velocity is strongly dominated by submesoscale turbulence.
 The hourly velocity fields are significantly
ageostrophic as depicted by the jPDF of vorticity-Laplacian of SSH
winter (Figure \ref{fig3}e). Even in April, only the daily-averaged fields
are largely
in  geostrophic balance (the jPDF of daily-averaged vorticity vs.\\ Laplacian of SSH
is an ellipse with large eccentricity and main axis tilted by 45$^\circ$; Figure \ref{fig3}f).

The October vorticity-strain jPDF shows much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
that persist in summer) and one with a very steep slope.
That the submesoscale dynamics in October are mainly unbalanced is clearly depicted
in the shape of jPDF of vorticity-Laplacian of SSH for hourly fields
 (Figure \ref{fig3}g), which is an ellipse aligned in the vertical axis.
Daily-averaging the model suppresses the ageostrophic, super-inertial flows, and, therefore,
the daily-averaged flow is essentially geostrophic as depicted by the 45$^\circ$-tilted
ellipse in the vorticity-Laplacian of SSH jPDF.

@@

To better quantify the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance.  Before calculating the spectra, time-mean and spatial linear
trends were removed, and the resulting fields were multiplied by a two-dimensional Hanning ``spectral window''.
The two-dimensional spectra were averaged azimuthaly \citep[e.g., ][]{rocha_etal2016} .
We discuss only spectra for the 1/48$^\circ$ simulation, which extends the 1/24$^\circ$ simulation
towards smaller scales (supplemental material).

\begin{figure*}[ht]
  \begin{center}
    \includegraphics[width=1.\textwidth]{figs/fig4.pdf}
 \caption{Surface (horizontal) KE (a) and SSH variance wavenumber spectra (b)
 in the 1/48$^\circ$ simulation. Solid lines
 are spectra based on hourly snapshots, and dashed lines are spectra based on daily-averaged
 fields.}
 \label{fig4}
 \end{center}
\end{figure*}

Figure \ref{fig4}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the April and October spectra based on hourly
velocity snapshots (solid lines) are nearly indistinguishable from each other within 95$\%$ confidence level.
Consistent with the results of
\citet{rocha_etal2016}, who analyzed a 3-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km, both in April and October (compare solid lines against dashed lines in Figure
\ref{fig4}a). But this suppression is dramatic in October, when the inertia-gravity waves peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

@@

An important caveat is that this study focuses on a single patch of ocean in the
vicinity of the Kuroshio Extension. This may be typical of a mesoscale-rich
subtropical region, but it is unlikely to be
representative of other regions such as low-eddy-kinetic-energy eastern boundary currents
and the middle of the subtropical gyre.  We plan to report on the geographic variability of submesoscale
 seasonality in a future study.

@@

% say something about relative energy in smoother vs.\\ unsmoothed fields.
% also say something about comparison with observations.

@@

\begin{acknowledgments}
 William R. Young provided helpful feedback on the first draft. We thank the MITgcm community and our colleagues at the NASA Advanced
Supercomputing (NAS) Division for their awesome support.
This research was funded by NSF (OCE1357047) and NASA (NNX13AE44G,NNX13AE85G,NNX16AH67G).
The LLC output can be obtained from the ECCO project (\texttt{http://ecco2.org/llc$\_$hires}). The altimeter products were produced by Ssalto/Duacs
and distributed by Aviso, with support from CNES (\texttt{http://www.aviso.altimetry.fr/duacs/}).
\end{acknowledgments}
